Covariate analysis of UFL-Mayo-ISB RNASeq (both cerebellum and temporal cortex)

### 2016-05-16 code/Rmd/Mayo_RNASeq_CellMarkersAdjusted.Rmd
Added markdown to adjust for cell types in Mayo RNASeq data

knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq.Rmd",
                                 parentId = "syn5570325",
                                 entityName = 'UFL_Mayo_ISB RNASeq')
```

@@

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
dev.off()
```

@@

thisFileName <- 'Mayo_RNASeq.Rmd'

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath")

@@

Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data1, fig.height=15, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E)

@@

pl = list()
p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Tissue), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
p <- p + geom_text(aes(label= ID), size=4, hjust=0)
pl[[1]] <- p

@@

WGCNA::plotDendroAndColors(sampleTrees, 
                           labels2colors(cbind(tree,COVARIATES.tmp)),
                           groupLabels = c('Cluster',colnames(COVARIATES.tmp)))
dev.off()

tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[tree == 0], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[tree == 3], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
```
### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, results='asis', echo=FALSE}
gc()
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, COVARIATES,
                                                 'NULL design(voom-normalized)',
                                                 isKeyPlot=TRUE)

@@

pl = list()
p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Tissue), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
p <- p + geom_text(aes(label= ID), size=4, hjust=0)
pl[[1]] <- p

@@

names(clust) = sampleTrees$labels
clust = clust[rownames(COVARIATES)]
```
```{r decompse.normalise.data4, fig.height=10, fig.width=25, results='asis'}
# Tree based clustering
COVARIATES.tmp = COVARIATES
COVARIATES.tmp[, FactorCovariates] = sapply(COVARIATES.tmp[,FactorCovariates], as.numeric)
plotDendroAndColors(sampleTrees, 
                    colors = labels2colors(cbind(data.frame(clust = clust), COVARIATES.tmp)), 
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES))
gc()
dev.off()

tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[tree == 0], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
tmp = paste('Samples in cluster 1 are', paste(sampleTrees$labels[tree == 1], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[tree == 3], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
```

@@

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB RNASeq', parentId = parentId)
CODE <- synStore(CODE)

>>>>

knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_CellMarkersAdjusted.Rmd",
                                 parentId = "syn5570325",
                                 entityName = 'UFL_Mayo_ISB RNASeq Cell Markers Adjusted')
```

@@

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

@@

thisFileName <- 'Mayo_RNASeq_CellMarkersAdjusted.Rmd'

@@

### Cell type markers (i.e., eigen genes of cell types) as additional covariates
```{r cellMark}
load(synGet('syn4893059')@filePath)
GeneSets = GeneSets[grep("Zhang", names(GeneSets))] %>%
  ldply(.fun = function(x){
    data.frame(hgnc_symbol = x)
  }, .id = 'CellType') %>%
  dplyr::mutate(CellType = gsub("Zhang:","",CellType)) %>%
  left_join(Ensemble2HGNC) %>%
  filter(!is.na(ensembl_gene_id))

kable(data.frame(table(GeneSets$CellType)) %>%
        dplyr::rename(CellType = Var1, NumberOfGenes = Freq))

COVARIATES.cellmarkers = dlply(GeneSets, .(CellType), .fun = function(gs){
  EXPR = COUNT[rownames(COUNT) %in% gs$ensembl_gene_id, ]
  PC <- prcomp(EXPR, scale.=T, center = T)
  rownameToFirstColumn(PC$rotation[,1, drop=F], 'ID') %>%
    plyr::rename(c('PC1' = unique(gs$CellType)))
}) %>%
  plyr::join_all(by = 'ID')

METADATA = join_all(list(METADATA, COVARIATES.cellmarkers))

p = METADATA %>%
  dplyr::select(Diagnosis, Tissue, one_of(colnames(COVARIATES.cellmarkers))) %>%
  gather(CellType, EigenGeneExpression, -ID, -Diagnosis, -Tissue) %>%
  ggplot(aes(x = Diagnosis, y = abs(EigenGeneExpression), fill = Tissue)) + geom_boxplot() + facet_grid(.~CellType)
p                         
```

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath", "Astrocyte", "Endothelial", "Microglia", "MyelinOligos", "Neuron", "NewOligos", "OPC")

@@

Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data1, fig.height=15, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E)

@@

pl = list()
p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Tissue), size=as.numeric(OPC)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
pl[[1]] <- p

p <- ggplot(plotdata, aes(x=PC3, y=PC4))
p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Diagnosis), size=as.numeric(OPC)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
pl[[2]] <- p

@@

WGCNA::plotDendroAndColors(sampleTrees, 
                           labels2colors(cbind(tree,COVARIATES.tmp[, FactorCovariates])),
                           groupLabels = c('Cluster',colnames(COVARIATES.tmp[, FactorCovariates])))
dev.off()

tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[tree == 0], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[tree == 3], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
```
### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, results='asis', echo=FALSE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, COVARIATES,
                                                 'NULL design(voom-normalized)',
                                                 isKeyPlot=TRUE)

@@

pl = list()
p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=factor(Source), color=as.factor(Tissue), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") +scale_shape_manual(values = 1:2)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
pl[[1]] <- p

@@

names(clust) = sampleTrees$labels
clust = clust[rownames(COVARIATES)]
```
```{r decompse.normalise.data4, fig.height=10, fig.width=25, results='asis'}
# Tree based clustering
COVARIATES.tmp = COVARIATES
COVARIATES.tmp[, FactorCovariates] = sapply(COVARIATES.tmp[,FactorCovariates], as.numeric)
plotDendroAndColors(sampleTrees, 
                    colors = labels2colors(cbind(data.frame(clust = clust), COVARIATES.tmp[,FactorCovariates])), 
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES))

tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[tree == 0], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[tree == 3], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
```

@@

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB RNASeq Cell Markers Adjusted', parentId = parentId)
CODE <- synStore(CODE)
